Reducing iptables configuration complexity using chains

### 2014-05-03 iptables/paper.tex
add section introducing chains

To make your rules more manageable, you can make use of chains in your iptables rules. This paper is inspired by an article that uses chains to make iptables more efficient (faster). The goal of this paper is to get iptables rules that are easier to read and configure, but it will result in faster handling of packets as well.\\
\section{Brief introduction to iptables}
\subsection{ESTABLISHED state}
When using this option, you can filter for established connections. If you define it in both the INPUT and OUTPUT rules, you only have to define in the INPUT rules which NEW incoming requests should be allowed, and in the OUTPUT rules which NEW outgoing request are allowed. The established connections will be allowed and should not be redefined (making the configuration a lot more readable and maintainable). An example allowing only an ssh service without using the ESTABLISHED state would be :
\begin{verbatim}
  iptables -A INPUT -p tcp --dport ssh -j ACCEPT
  iptables -A INPUT -j REJECT
  iptables -A OUTPUT -p tcp --sport ssh -j ACCEPT
  iptables -A OUTPUT -j REJECT}
\end{verbatim}

@@

In this example, the benefit of using the connection state is not clear, but when more allowed incoming services are added, they only have to be added on the INPUT chain, but not on the OUTPUT chain, because they are covered by the ESTABLISHED rule.\\
In the first example (without the ESTABLISHED rule), every allowed incoming connection should be repeated in the OUTPUT chain, matching the packets sent for the outgoing connection, which results in an equal amount of rules on both chains.\\
If you want to do filtering in both directions (allowing incoming request for listening services and outgoing request for remote services), this can become very messy, and almost unmaintainable without making mistakes.\\
\section{Optimizing configuration with an example setup}
\section{Conclusion}

>>>>

To make your rules more manageable, you can make use of chains in your iptables rules. This paper is inspired by an article that uses chains to make iptables more efficient (faster). The goal of this paper is to get iptables rules that are easier to read and configure, but it will result in faster handling of packets as well.\\

\section{Brief introduction to iptables}
\subsection{ESTABLISHED state}
When using this option, you can filter for established connections. If you define it in both the INPUT and OUTPUT rules, you only have to define in the INPUT rules which NEW incoming requests should be allowed, and in the OUTPUT rules which NEW outgoing request are allowed. The established connections will be allowed and should not be redefined (making the configuration a lot more readable and maintainable). An example allowing only an ssh service without using the ESTABLISHED state would be :
\begin{verbatim}
  iptables -A INPUT -p tcp --dport ssh -j ACCEPT
  iptables -A INPUT -j REJECT
  iptables -A OUTPUT -p tcp --sport ssh -j ACCEPT
  iptables -A OUTPUT -j REJECT}
\end{verbatim}

@@

In this example, the benefit of using the connection state is not clear, but when more allowed incoming services are added, they only have to be added on the INPUT chain, but not on the OUTPUT chain, because they are covered by the ESTABLISHED rule.\\
In the first example (without the ESTABLISHED rule), every allowed incoming connection should be repeated in the OUTPUT chain, matching the packets sent for the outgoing connection, which results in an equal amount of rules on both chains.\\
If you want to do filtering in both directions (allowing incoming request for listening services and outgoing request for remote services), this can become very messy, and almost unmaintainable without making mistakes.\\

\subsection{Introducing chains}
When two services (on different ports) should be available to a limited but identical list of IP addresses.
Without using chains, for every combination of port and IP a rule should be created :
\begin{verbatim}
  iptables -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 22 -j ACCEPT
  iptables -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 22 -j ACCEPT
  iptables -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 22 -j ACCEPT

  iptables -A INPUT -p tcp -m tcp -s 10.100.2.3 --dport 10000 -j ACCEPT
  iptables -A INPUT -p tcp -m tcp -s 10.100.2.4 --dport 10000 -j ACCEPT
  iptables -A INPUT -p tcp -m tcp -s 10.100.2.7 --dport 10000 -j ACCEPT
\end{verbatim}

Resulting in a lot of rules, and when an IP address has to be changed, added or removed, this has to be done for every corresponding rule.\\

When using chains, this can be much easier. Imagine, that you first check if the packet matches the destination port, and if it does, jump to a new chain, where a list of IP addresses is checked. :
\begin{verbatim}
  // create new chain admin_IP
  iptables -N admin_IP

  // add rules to chain admin_IP
  iptables -A admin_IP -s 10.100.2.3 -j ACCEPT
  iptables -A admin_IP -s 10.100.2.4 -j ACCEPT
  iptables -A admin_IP -s 10.100.2.7 -j ACCEPT
  // drop all packets that are not matched by previous rules
  iptables -A admin_IP -j DROP

  // filter ports in INPUT chain
  iptables -A INPUT -p tcp -m tcp --dport 22 -j admin_IP
  iptables -A INPUT -p tcp -m tcp --dport 10000 -j admin_IP
\end{verbatim}

As you can see, there is are several benefits of putting the IP addresses in a separate chain :
\begin{itemize}
    \item the list of IP addresses in the separate chain can be reused for both ports, so they have to be defined only once.
    \item adding/changing/removing an IP address is much easier
    \item there is a better overview of the firewall rules
\end{itemize}

\section{Optimizing configuration with an example setup}
\section{Conclusion}
